---
title: "GWAS_introduction_inspection"
author: "Marc Vaudel"
date: '2022-05-23'
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Inspecting GWAS summary statistics

This notebook provides and introduction to inspecting the results of genome-wide association studies.

## Libraries

We will need the following libraries. If they are not availble on your system, use the `ìnstall_packages` command to install them.

```{r }

library(janitor) # A package to clean dirty data
library(tidyr) # A package to tidy data
library(dplyr) # A package to manipulate data
library(parallel) # A package to run tasks in parallel
library(glue) # A package to glue variables in strings

library(curl) # A package to connect to urls
library(httr) # A package to run http queries
library(rjson) # A package to handle json files

library(LDlinkR) # A package to query LDlink
library(phenoscanner) # A package to query phenoscanner  NOTE: phenoscanner needs to be installed from https://github.com/phenoscanner/phenoscanner

library(ggplot2) # A package to plot data
library(scico) # A package of scientific color palettes
library(RColorBrewer) # A package of color palettes

library(conflicted) # A package to manage namespace conflicts

# Resolve conflicting function names
conflict_prefer("filter", "dplyr")

# Set the theme for plots
theme_set(theme_bw(base_size = 13))

# Ensembl rest APIs
server <- "https://rest.ensembl.org"
server37 <- "https://grch37.rest.ensembl.org"


```

## Inspecting GWAS summary statistics

Load results from the birth weight GWAS available at this [link](https://drive.google.com/file/d/1mn64G91258pDdDARVAS1cYlcof-Knu2n/view?usp=sharing).

```{r }

file_path <- "/home/marc/Projects/tutorials/data/Fetal_BW_European_meta.NG2019.txt.gz"

gwas_results <- read.table(
  file = file_path,
  header = T,
  sep = " ",
  stringsAsFactors = F
) %>% 
  clean_names()

```

_Do you understand the meaning of the different columns? Is there anything that you find surprising?_

Build a Manhattan plot. Note: there are dedicated packages that can do this but for the sake of the exercise we will write down the code.

```{r }

# Chromosome lengths in GRCh37.p13 (hg19) from Ensembl

chromosomes <- c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11", "12", "13", "14", "15", "16", "17", "18", "19", "20", "21", "22", "X")
chromosomeLength <- c(249250621, 243199373, 198022430, 191154276, 180915260, 171115067, 159138663, 146364022, 141213431, 135534747,	135006516, 133851895, 115169878, 107349540, 102531392, 90354753, 81195210, 78077248, 59128983, 63025520, 48129895, 51304566, 	155270560)
genomeLength <- sum(chromosomeLength)
chromosomeStart <- cumsum(chromosomeLength) - chromosomeLength
chromosomeMiddle <- chromosomeStart + chromosomeLength / 2

# Arrange for plotting

plot_data <- gwas_results %>%
  mutate(
    logP = -log10(p)
  ) %>% 
  arrange(
    logP
  ) %>%
  mutate(
    chromosomeNumber = as.numeric(ifelse(chr == 'X', 23, chr)),
    x = chromosomeStart[chromosomeNumber] + pos,
    color = factor(chromosomeNumber %% 2, levels = 0:1)
  ) %>%
  arrange(
    chromosomeNumber, logP, pos
  )

maxP <- max(plot_data$logP)

# Colors

mhColor1 <- "grey20"
mhColor2 <- "grey40"
colors <- c(mhColor1, mhColor2)


# Chromosome labels

xLabels <- 1:22
xLabels[xLabels %% 2 == 0 & xLabels > 17] <- ""
xLabels <- c(xLabels, "X")

# y axis

maxY <- 5 * ceiling(max(plot_data$logP / 5))
maxY <- max(maxY, 10)

yBreaks <- c(0, 5, -log10(5e-8))
yLabels <- c("", 5, round(-log10(5e-8), digits = 1))

lastBreak <- floor(max(maxY / 10))

if (lastBreak > 0) {
  
  newBreaks <- 10*(1:lastBreak)
  
  while(length(newBreaks) > 3) {
    
    newBreaks <- newBreaks[c(T, F)]
    
  }
  
  yBreaks <- c(yBreaks, newBreaks)
  yLabels <- c(yLabels, round(newBreaks, digits = 1))
  
}


# Build plot
ggplot(
  data = plot_data
) + 
  geom_hline(
    yintercept = -log10(5e-8), 
    col = "green4", 
    size = 0.3
  ) + 
  geom_point(
    aes(
      x = x, 
      y = logP, 
      col = color
    ), 
    size = 1
  ) +
  scale_y_continuous(
    name = "p-value [-log10]", 
    breaks = yBreaks, 
    labels = yLabels, 
    expand = expansion(
      mult = c(0, 0.05)
    ), 
    limits = c(0, maxY)
  ) + 
  scale_x_continuous(
    name = "Chromosome", 
    breaks = chromosomeMiddle, 
    labels = xLabels, 
    limits = c(0, genomeLength), 
    expand = expansion(
      mult = 0.01
    )
  ) + 
  scale_color_manual(
    values = colors
  ) + 
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(size = 0.3),
    strip.background = element_rect(
      fill = "grey99"
    )
  )

```

Build a QQ plot. Note: there are dedicated packages that can do this but for the sake of the exercise we will write down the code.

```{r }

# Compute expected p-values

plot_data <- gwas_results %>%
  mutate(
    logP = -log10(p)
  ) %>% 
  arrange(
    logP
  ) %>% 
  mutate(
    expectedLogP = sort(-log10(ppoints(n = n())))
  )

# Axis

maxValue <- max(plot_data$logP, plot_data$expectedLogP, 10)


# Build plot
ggplot(
  data = plot_data
) + 
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dotted"
  ) +
  geom_point(
    mapping = aes(
      x = expectedLogP,
      y = logP
    ),
    size = 1
  ) +
  scale_y_continuous(
    name = "p-value [-log10]", 
    limits = c(0, maxValue),
    expand = expansion(
      mult = 0.02
    )
  ) +
  scale_x_continuous(
    name = "Expected p-value [-log10]",
    limits = c(0, maxValue),
    expand = expansion(
      mult = 0.02
    )
  )

```

Build a histogram of allele frequencies.

```{r }

# Compute the maf
plot_data <- plot_data %>% 
  mutate(
    minor_allele_frequency = ifelse(eaf > 0.5, 1 - eaf, eaf)
  )

# Build histogram
ggplot(
  data = plot_data
) + 
  geom_histogram(
    mapping = aes(
      x = 100 * minor_allele_frequency,
    ),
    size = 1,
    bins = 50
  ) +
  scale_y_continuous(
    name = "# Variants",
    expand = expansion(
      mult = c(0, 0.05)
    )
  ) +
  scale_x_continuous(
    name = "Effect allele frequency [%]"
  )

```

This study included variants down to a minor allele frequency of `r 100*min(plot_data$minor_allele_frequency)` %.

Build some maf categories.

```{r }

# Compute the maf
plot_data <- plot_data %>% 
  mutate(
    maf_category = case_when(
      minor_allele_frequency < 0.005 ~ "< 0.5 %",
      minor_allele_frequency < 0.05 ~ "< 5 %",
      T ~ ">= 5%"
    ),
    maf_category = factor(maf_category, levels = c("< 0.5 %", "< 5 %", ">= 5%"))
  )

table(plot_data$maf_category)/nrow(plot_data)

```

Build a QQ plot stratifying the variants per maf category.

```{r }

# Compute expected p-values

plot_data_stratified <- plot_data %>%
  group_by(
    maf_category
  ) %>% 
  arrange(
    logP
  ) %>% 
  mutate(
    expectedLogP = sort(-log10(ppoints(n = n())))
  ) %>% 
  ungroup()

# Axis

maxValue <- max(plot_data_stratified$logP, plot_data_stratified$expectedLogP, 10)


# Build plot
ggplot(
  data = plot_data_stratified
) + 
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dotted"
  ) +
  geom_point(
    mapping = aes(
      x = expectedLogP,
      y = logP,
      col = maf_category
    ),
    size = 1
  ) +
  scale_y_continuous(
    name = "p-value [-log10]", 
    limits = c(0, maxValue),
    expand = expansion(
      mult = 0.02
    )
  ) +
  scale_x_continuous(
    name = "Expected p-value [-log10]",
    limits = c(0, maxValue),
    expand = expansion(
      mult = 0.02
    )
  )

```

Plot the effect size estimate against the effect allele frequency.

```{r }

# Color the genome-wide significant hits differently

plot_data <- plot_data %>% 
  mutate(
    gwas_significant = ifelse(p <= 5e-8, "p <= 5e-8", "p > 5e-8"),
    gwas_significant = factor(gwas_significant, levels = c("p <= 5e-8", "p > 5e-8"))
  )

# Build plot
ggplot(
  data = plot_data
) + 
  geom_hline(
    yintercept = 0
  ) +
  geom_point(
    mapping = aes(
      x = 100 * eaf,
      y = beta,
      col = gwas_significant
    ),
    size = 1
  ) +
  scale_y_continuous(
    name = "Effect size estimate"
  ) +
  scale_x_continuous(
    name = "Effect allele frequency [%]"
  ) +
  scale_color_manual(
    name = "Significance",
    values = c("darkred", "grey")
  )

```

## Zooming in on a specfic signal

We will now focus on the region around 123 Mbp.

```{r }

region_data <- read.table(
  file = "resources/region_data.gz",
  header = T,
  sep = "\t",
  stringsAsFactors = F
)

# Build plot
ggplot(
  data = region_data
) + 
  geom_point(
    mapping = aes(
      x = pos/10e6,
      y = -log10(p)
    ),
    size = 1
  ) +
  scale_x_continuous(
    name = "Position [Mbp]",
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    name = "P-value [-log10]",
    expand = expansion(
      mult = c(0, 0.05)
    )
  ) + 
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )

```

Find the rsid of the variant with best p-value in that region.

```{r }

best_rsid <- region_data$rsid[which.min(region_data$p)]

```

_Look for the variants in LD with this SNP using LDlink - how would you do this for hundreds of variants?_

Find the variants in LD with this variant using LDlink.

```{r }

window <- "500000"
token <- "b823fa70d49d" # Note: you should get your own authentification token from LDlink

ld_resutls <- LDproxy(
  snp = best_rsid, 
  pop = "CEU", 
  r2d = "r2", 
  token = token
) %>% 
  clean_names()

```

Annotate the LD with the lead SNP on the plot and add recombination rates.

```{r }

# Merge LD values

r2_values <- ld_resutls %>% 
  select(
    rsid = rs_number, ld = r2
  )

plot_data <- region_data %>% 
  left_join(
    r2_values,
    by = "rsid"
  ) %>% 
  mutate(
    ld = ifelse(is.na(ld), 0, ld),
    ldFactor = case_when(
      ld < 0.2 ~ "[0.0  0.2]",
      ld >= 0.2 & ld < 0.4 ~ "[0.2  0.4]",
      ld >= 0.4 & ld < 0.6 ~ "[0.4  0.6]",
      ld >= 0.6 & ld < 0.8 ~ "[0.6  0.8]",
      ld >= 0.8 ~ "[0.8  1.0]"
    ),
    ldFactor = factor(ldFactor, c("[0.0  0.2]", "[0.2  0.4]", "[0.4  0.6]", "[0.6  0.8]", "[0.8  1.0]"))
  ) %>% 
    arrange(
        ldFactor
    )

lead_snp <- plot_data %>% 
  filter(
    rsid == best_rsid
  )

bestP <- max(-log10(lead_snp$p))

# Get recombination rates

recombination_rates <- read.table(
  file = "resources/recombination_rates/genetic_map_GRCh37_chr3.txt.gz",
  header = T,
  sep = "\t",
  stringsAsFactors = F
) %>% 
  clean_names() %>% 
  filter(
    position_bp >= min(plot_data$pos) & position_bp <= max(plot_data$pos)
  ) %>% 
  mutate(
    recombination_rate_scaled = rate_c_m_mb / max(rate_c_m_mb) * bestP * 2 / 3
  )

# Build plot
ggplot() + 
  geom_line(
    data = recombination_rates,
    mapping = aes(
      x = position_bp,
      y = recombination_rate_scaled
    ),
    col = "blue3"
  ) +
  geom_point(
  data = plot_data,
    mapping = aes(
      x = pos,
      y = -log10(p),
      col = ldFactor
    ),
    size = 1
  ) +
    geom_point(
        data = lead_snp,
        mapping = aes(
      x = pos,
      y = -log10(p)
        ),
        shape = 18,
        col = "black",
        size = 4
    ) +
    geom_point(
        data = lead_snp,
        mapping = aes(
      x = pos,
      y = -log10(p)
        ),
        shape = 18,
        col = "red3",
        size = 3
    ) +
  scale_x_continuous(
    name = "Position",
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    name = "P-value [-log10]",
    expand = expansion(
      mult = c(0, 0.05)
    )
  ) +
    scale_color_manual(
        name = "R2",
        values = brewer.pal(9, "Set1")[c(9, 2, 3, 5, 1)],
        guide = guide_legend(
            override.aes = list(
                size = 4
            )
        )
    )  + 
  theme(
    legend.position = "top",
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )

```

_Look for the possible consequences of this variant in Ensembl - how would you do this for hundreds of variants?_

Do this with the Ensembl API.

```{r }

ext <- "/vep/human/id"
body <- paste0('{ "ids" : ["', best_rsid, '" ] }') # Note that you can query multiple variants, just separate their ids with commas

r <- POST(
  url = paste(server37, ext, sep = ""), 
  content_type("application/json"), 
  accept("application/json"), 
  body = body
)

response_code <- status_code(r)

if (response_code != 200) {
  
  stop(paste0("Ensembl query failed for ", best_rsid, ". Status: ", responseCode, "."))
  
}

vep_results <- fromJSON(toJSON(content(r)))

```

_Do you find the possible consequences?_

Similarly, query the genes in this region.

```{r }

chromosome <- 3
position <- region_data$pos[region_data$rsid == best_rsid]
position_start <- max(position - 500000, 0)
position_end <- position + 500000

region <- paste0(chromosome, ":", position_start, "-", position_end)

ext <- paste0("/overlap/region/human/", region, "?feature=gene")

r <- GET(
  url = paste(server37, ext, sep = ""), 
  content_type("application/json"), 
  accept("application/json")
)

response_code <- status_code(r)

if (response_code != 200) {
  
  stop(paste0("Ensembl query failed for ", best_rsid, ". Status: ", responseCode, "."))
  
}

gene_results <- fromJSON(toJSON(content(r)))

```

_What is the nearest gene?_

Now, see whether this variant was associated with other traits using phenoscanner.

```{r }

phenoscanner_results <- phenoscanner(
  snpquery = best_rsid,
  proxies = "EUR",
  pvalue = 5e-8
)

View(phenoscanner_results$results)

```

_How many other traits are associated with this variant?_


