---
title: "GWAS_introduction_association"
author: "Marc Vaudel"
date: '2022-05-23'
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## GWAS data introduction and QC

This notebook provides and introduction to conducting genome-wide association studies.

## Libraries

We will need the following libraries. If they are not availble on your system, use the `ìnstall_packages` command to install them.

```{r }

library(janitor) # A package to clean dirty data
library(tidyr) # A package to tidy data
library(dplyr) # A package to manipulate data
library(parallel) # A package to run tasks in parallel
library(glue) # A package to glue variables in strings

library(ggplot2) # A package to plot data

library(conflicted) # A package to manage namespace conflicts

# Resolve conflicting function names
conflict_prefer("filter", "dplyr")

# Set the theme for plots
theme_set(theme_bw(base_size = 13))

```


## Load data

Load the example data available at this [link](https://drive.google.com/file/d/1O1rBQ1ldOEyCp1yzh2w9hIkGFCrkRZlN/view?usp=drive_link).

```{r }

file_path <- "/home/marc/Projects/tutorials/data/lm_example.gz"

start_time <- Sys.time()

snp_data <- read.table(
  file = file_path,
  sep = "\t",
  header = T,
  stringsAsFactors = F
) %>% 
  clean_names()

end_time <- Sys.time()

processing_time <- difftime(end_time, start_time)

phenotypes <- c("pheno_a", "pheno_b", "pheno_c", "pheno_d")

```

The file was imported in `r round(processing_time[[1]], 1)` `r units(processing_time)``. 

_How can we reduce the time spent reading files?_

This data set, courtesy of Dr. Bram Burger, contains simulated genotypes and phenotypes as a table:
- chromosome: the chromosome where the variant is located
- position: the position of the variant, here we have only one variant per position
- genotype: the number of alternative alleles for a given variant in a given individual
- sample: the identifier of the individual
- pheno_x: the phenotypic values

There is one variant at every genomic position, making `r length(unique(snp_data$position))` variants for `r length(unique(snp_data$sample))` individuals.

_How does this compare to GWAS studies you might have seen in the literature? How long would it take to load a similar file containing ten million variants?_

## Genotypes

Select a position, for the variant at this position, compute the number of participants with a given genotype.

```{r }

sampled_position <- sample(snp_data$position, 1)

pos_data <- snp_data %>% 
  filter(
    position == sampled_position
  )

print(
  glue("# Number of participants for a given genotype for the variant at position {sampled_position}")
)

table(pos_data$genotype)

```

Estimate the alternative allele frequency: `r round(100 * sum(pos_data$genotype) / (2 * nrow(pos_data)), 1)` %.

Load summary information on the variants in one chromosome of MoBa available at this [link](https://drive.google.com/file/d/18XqRm2PzI1JIUKB75ePw8aHV4CfNeGTR/view?usp=drive_link).

```{r }

file_path <- "/home/marc/Projects/tutorials/data/22-markerinfo.gz"

marker_info <- read.table(
  file = file_path,
  sep = "\t",
  header = F,
  stringsAsFactors = F
) %>% 
  clean_names()

names(marker_info) <- c("chromosome", "position", "id", "ref", "alt", "typed", "info", "ref_panel_af")

```

_What is the content of the different columns?_

Compute the share of genotyped markers.

```{r }

n_genotyped <- sum(marker_info$typed == 1)

total <- nrow(marker_info)

share_genotyped <- round(100 * n_genotyped / total, 1)

print(glue("{n_genotyped} markers of {total} are genotyped ({share_genotyped} %)"))

```

_How can we conduct genome-wide analyses when probing only `r share_genotyped` variants? How will this influence our results?_

Build a histogram of info scores for non-genotyped markers.

```{r }

non_genotyped_marker_info <- marker_info[marker_info$typed != 1, ]

ggplot() +
  geom_histogram(
    data = non_genotyped_marker_info,
    mapping = aes(
      x = info
    ),
    col = "darkred",
    fill = "darkred",
    alpha = 0.1,
    bins = 20
  ) +
  scale_x_continuous(
    name = "Info score"
  ) +
  scale_y_continuous(
    name = "# variants",
    expand = expansion(
      mult = c(0, 0.05)
    )
  )

```

_How would you use this variable to remove low quality markers?_

Compute the minor allele frequency in the reference panel, and build a histogram.

```{r }

non_genotyped_marker_info <- non_genotyped_marker_info %>% 
  mutate(
    minor_allele_frequency = 100 * ifelse(ref_panel_af > 0.5, 1 - ref_panel_af, ref_panel_af)
  )

ggplot() +
  geom_histogram(
    data = non_genotyped_marker_info,
    mapping = aes(
      x = minor_allele_frequency
    ),
    col = "darkblue",
    fill = "darkblue",
    alpha = 0.1,
    bins = 20
  ) +
  scale_x_continuous(
    name = "Minor allele frequency [%]"
  ) +
  scale_y_continuous(
    name = "# variants",
    expand = expansion(
      mult = c(0, 0.05)
    )
  )

```

_How does this compare to the histogram obtained previously?_

Plot the info score against the minor allele frequency.

```{r }

ggplot() +
  geom_point(
    data = non_genotyped_marker_info,
    mapping = aes(
      x = minor_allele_frequency,
      y = info
    ),
    alpha = 0.1
  ) +
  scale_x_continuous(
    name = "Minor allele frequency [%]"
  ) +
  scale_y_continuous(
    name = "Info score"
  )

```

_Can you explain this trend? Which markers would you keep for a GWAS?_

## Phenotypes

_What kind of variables can be used for phenotypes in GWASes?_

Build a histogram of the different phenotypes available in this data set.

```{r }

pheno_data <- pos_data %>% # The phenotypes are the same for all variants, we just need the data from one of them
  pivot_longer(
    cols = starts_with("pheno_"),
    names_prefix = "pheno_"
  ) %>% 
  select(
    sample, pheno = name, value
  )

ggplot() +
  geom_histogram(
    data = pheno_data,
    mapping = aes(
      x = value
    ),
    col = "darkgreen",
    fill = "darkgreen",
    alpha = 0.1,
    bins = 20
  ) +
  scale_x_continuous(
    name = "Phenotype value",
    breaks = c(-5:5)
  ) +
  scale_y_continuous(
    name = "# participants",
    expand = expansion(
      mult = c(0, 0.05)
    )
  ) + 
  facet_wrap(
    ~ pheno
  )

```

_How do you interpret these distributions? How would you process phenotypes prior to a GWAS?_

## Genotype to phenotype association

For the variant at the position selected below, look at the association between `genotype` and `pheno_a`.

```{r }

association_results <- lm(
  data = pos_data,
  formula = pheno_a ~ genotype
)

print(
  glue("# Association results at position {sampled_position}")
)

summary(association_results)

```

_How do you interpret these results? Is there a significant association?_

Plot the phenotypes against genotypes at this position.

```{r }

ggplot() +
  theme_bw() +
  geom_hline(
    yintercept = 0
  ) +
  geom_point(
    data = pos_data,
    mapping = aes(
      x = genotype,
      y = pheno_a
    ),
    alpha = 0.1
  ) +
  geom_smooth(
    data = pos_data,
    mapping = aes(
      x = genotype,
      y = pheno_a
    ),
    method = "lm",
    formula = "y ~ x"
  ) +
  scale_x_continuous(
    name = "Genotype",
    breaks = c(0, 1, 2)
  ) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )

```

_How would you expect the plot to look like for a significant association?_

Run the association for all variants and all phenotypes.

```{r }

#' Run a linear model for the variant at the given position against the given phenotype.
#' 
#' @param position The position of the variant.
#' @param snp_data The data frame containing the genotypes and phenotypes.
#' @return The coefficients of the regression as a data frame.
lm_function <- function(position, snp_data, pheno) {
  
  variant_data <- snp_data[snp_data$pos == position, ]
  
  lm_results <- lm(
    formula = glue("{pheno} ~ genotype"),
    data = variant_data
  )
  
  lm_summary <- summary(lm_results)
  
  coefficients <- data.frame(
    position = position,
    phenotype = pheno,
    beta = lm_summary$coef["genotype", "Estimate"],
    se = lm_summary$coef["genotype", "Std. Error"],
    p = lm_summary$coef["genotype", "Pr(>|t|)"],
    n = nrow(variant_data),
    stringsAsFactors = F
  )
  
  
  return(coefficients)
  
}

# Calculate the number of cores
no_cores <- max(detectCores() - 1, 1)

# Initiate a cluster to run functions on multiple cores
cluster <- makeCluster(no_cores)

clusterExport(cluster, "snp_data")
clusterExport(cluster, "lm_function")
clusterExport(cluster, "glue")

# Process phenotypes one after the other

start_time <- Sys.time()

lm_results <- NULL

for (phenotype in phenotypes) {
  
  print(glue("{Sys.time()} - Processing {phenotype}"))
  
  clusterExport(cluster, "phenotype")
  
  # Process variants in parallel
  
  pheno_results <- parLapply(
    cl = cluster, 
    X = sort(unique(snp_data$pos)), 
    fun = function (position) lm_function(position, snp_data, phenotype) 
  )
  
  # Merge into a single table
  
  pheno_results <- do.call("rbind", pheno_results)
  
  lm_results <- rbind(lm_results, pheno_results)
  
}

end_time = Sys.time()

stopCluster(cluster)

processing_time = difftime(end_time, start_time)

```

The analysis of these `r length(unique(snp_data$position))` variants and `r length(phenotypes)` phenotypes using  `r no_cores` threads was conducted in `r round(processing_time[[1]], 1)` `r units(processing_time)`. 

_How long would it take to process one phenotype over the entire genome? How can we speed up this process?_

Plot the p-value against location for all phenotypes.

```{r }

lm_results$phenoFactor <- factor((lm_results$phenotype), levels = phenotypes)

levels(lm_results$phenoFactor) <- c("a", "b", "c", "d")

ggplot() +
  geom_point(
    data = lm_results,
    mapping = aes(
      x = position,
      y = -log10(p)
    )
  ) +
  scale_x_continuous(
    "Position [bp]"
  ) +
  scale_y_continuous(
    name = "p-value [-log10]",
    expand = expansion(
      mult = c(0, 0.1)
    )
  ) +
  facet_grid(
    phenotype ~ .
  )

```

Compute and plot the association of phenotype b with genotypes for the variant of strongest association.

```{r }

best_pos <- lm_results$position[which.max(-log10(lm_results$p))]

best_pos_data <- snp_data %>% 
  filter(
    position == best_pos
  )

association_results <- lm(
  data = best_pos_data,
  formula = pheno_b ~ genotype
)

print(
  glue("# Association results at position {sampled_position}")
)

summary(association_results)

```

```{r }

ggplot() +
  theme_bw() +
  geom_hline(
    yintercept = 0
  ) +
  geom_point(
    data = pos_data,
    mapping = aes(
      x = genotype,
      y = pheno_b
    ),
    alpha = 0.1
  ) +
  geom_smooth(
    data = pos_data,
    mapping = aes(
      x = genotype,
      y = pheno_b
    ),
    method = "lm",
    formula = "y ~ x"
  ) +
  scale_x_continuous(
    name = "Genotype",
    breaks = c(0, 1, 2)
  ) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )

```

_What do you think about the strength of the association? How can we reach genome-wide significance?_


## Filter by minor allele count

Remove variants with a minor allele count (mac) below 200.

```{r }

mac_threshold <- 200

mac <- snp_data %>% 
  group_by(
    position
  ) %>% 
  summarize(
    mac = ifelse(sum(genotype) > n(), 2 * n() - sum(genotype), sum(genotype))
  )

variants_excluded <- mac$position[mac$mac < mac_threshold]
variants_kept <- mac$position[mac$mac >= mac_threshold]

print(glue(("{length(variants_excluded)} variants excluded ({100 * length(variants_excluded)/(length(variants_kept) + length(variants_excluded))} %) using a mac threshold of {mac_threshold}.")))

lm_results <- lm_results[lm_results$position %in% variants_kept, ]

```

_Why filter out these variants? Why use a minor allele count (mac) intead of a minor allele frequency (maf) threshold?_

> Note that this step is usually done prior to running the GWAS. It is important to log the variants excluded.

Plot again the p-value per position.

```{r }

ggplot() +
  geom_point(
    data = lm_results,
    mapping = aes(
      x = position,
      y = -log10(p)
    )
  ) +
  scale_x_continuous(
    "Position [bp]"
  ) +
  scale_y_continuous(
    name = "p-value [-log10]",
    expand = expansion(
      mult = c(0, 0.1)
    )
  ) +
  facet_grid(
    phenotype ~ .
  )

```

_Do you notice any difference? What other variable can we use to filter and reduce the prevalence of spurious annotation?_


## Accounting for batches, covariates, and admixture

The genotypes we acquired and QCed in different batches. Our phenotypes are known to be associated with age and sex, with differences between males and females. Also, our population presents some level of admixture.

Convert the sex as number

```{r }

snp_data$sex_number <- as.numeric(factor(snp_data$sex, levels = c("M", "F")))

print(glue("Prevalence of sexes in percent:"))
100 * table(snp_data$sex_number)/nrow(snp_data)

```

_Can we do the same for batches?_

```{r }

print(glue("Prevalence of batches in percent:"))
100 * table(snp_data$batch)/nrow(snp_data)

```

Convert the batch value using one-hot encoding.

```{r }

for(batch in unique(snp_data$batch)) {
  
  snp_data[[batch]] <- ifelse(snp_data$batch == batch, 1, 0)
  
  print(glue("Prevalence of {batch}: {100 * sum(snp_data[[batch]])/nrow(snp_data)} %"))
  
}

```

Include age, sex, batches, and ten PCs to the model and rerun.

```{r }

#' Run a linear model for the variant at the given position against the given phenotype.
#' 
#' @param position The position of the variant.
#' @param snp_data The data frame containing the genotypes and phenotypes.
#' @return The coefficients of the regression as a data frame.
lm_function <- function(position, snp_data, pheno) {
  
  variant_data <- snp_data[snp_data$pos == position, ]
  
  lm_results <- lm(
    formula = glue("{pheno} ~ genotype + sex_number + age + sex_number * age + batch_1 + batch_2 + batch_3 + pc1 + pc2 + pc3 + pc4 + pc5 + pc6 + pc7 + pc8 + pc9 + pc10"),
    data = variant_data
  )
  
  lm_summary <- summary(lm_results)
  
  coefficients <- data.frame(
    position = position,
    phenotype = pheno,
    beta = lm_summary$coef["genotype", "Estimate"],
    se = lm_summary$coef["genotype", "Std. Error"],
    p = lm_summary$coef["genotype", "Pr(>|t|)"],
    n = nrow(variant_data),
    stringsAsFactors = F
  )
  
  
  return(coefficients)
  
}

# Calculate the number of cores
no_cores <- max(detectCores() - 1, 1)

# Initiate a cluster to run functions on multiple cores
cluster <- makeCluster(no_cores)

clusterExport(cluster, "snp_data")
clusterExport(cluster, "lm_function")
clusterExport(cluster, "glue")

# Process phenotypes one after the other

start_time <- Sys.time()

lm_results <- NULL

for (phenotype in phenotypes) {
  
  print(glue("{Sys.time()} - Processing {phenotype}"))
  
  clusterExport(cluster, "phenotype")
  
  # Process variants in parallel
  
  pheno_results <- parLapply(
    cl = cluster, 
    X = sort(unique(snp_data$pos)), 
    fun = function (position) lm_function(position, snp_data, phenotype) 
  )
  
  # Merge into a single table
  
  pheno_results <- do.call("rbind", pheno_results)
  
  lm_results <- rbind(lm_results, pheno_results)
  
}

end_time = Sys.time()

stopCluster(cluster)

processing_time = difftime(end_time, start_time)

```

The analysis of these `r length(unique(snp_data$position))` variants and `r length(phenotypes)` phenotypes using  `r no_cores` threads was conducted in `r round(processing_time[[1]], 1)` `r units(processing_time)`.

Plot again the p-value per position.

```{r }

ggplot() +
  geom_point(
    data = lm_results,
    mapping = aes(
      x = position,
      y = -log10(p)
    )
  ) +
  scale_x_continuous(
    "Position [bp]"
  ) +
  scale_y_continuous(
    name = "p-value [-log10]",
    expand = expansion(
      mult = c(0, 0.1)
    )
  ) +
  facet_grid(
    phenotype ~ .
  )

```
Look at the association summary stats for the variant of strongest association.

```{r }

best_pos <- lm_results$position[which.max(-log10(lm_results$p))]

best_pos_data <- snp_data %>% 
  filter(
    position == best_pos
  )

association_results <- lm(
  data = best_pos_data,
  formula = pheno_b ~ genotype + sex_number + age + sex_number * age + batch_1 + batch_2 + batch_3 + pc1 + pc2 + pc3 + pc4 + pc5 + pc6 + pc7 + pc8 + pc9 + pc10
)

print(
  glue("# Association results at position {sampled_position}")
)

summary(association_results)

```

_Do you notice any difference?_

## Running a GWAS taking into account relatedness

Multiple tools allow running a GWAS. An example of script using [regenie](rgcgithub.github.io/regenie/) is available at [resources/run_regenie.snake](resources/run_regenie.snake).

_What are the different rules? Do you recognize some options from the above example?_

