---
title: "Analysis of Quantitative Proteomics Data"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The result of a proteomic quantitative analysis is a list of peptide and protein abundances for every protein in different samples, or abundance ratios between the samples. The downstream interpretation of these data varies according to the experimental design. In this chapter we will describe different generic methods for the interpretation of quantitative datasets.

The dataset used here for illustrative purposes is freely available through the ProteomeXchange [(1)](#references) consortium via the PRIDE [(2)](#references) partner repository under the accession number PXD000441, and available in the _resources_ folder under _Results_MQ_5cell-line-mix_. It consists of five cell lines derived from acute myeloid leukemia (AML) patients measured with a spiked-in internal stanard (IS) obtained from the combination of the same five AML cell lines that have been metabolically labeled with heavy isotopes [(3)](#references) and analyzed using MaxQuant [(4)](#references) version 1.4.1.2. See [(5)](#references) for details.

This chapter introduces the basic methods, and is by no means aiming at covering all possibilities or present a reference workflow. Please continue exploring the data on your own and critically adapt the interpretation workflow to your experiment - there is no one workflow fits all!

## Installation

This tutorial uses [R](https://www.r-project.org), a language that allows the simple manipulation of large datasets. We will use R from the open source [RSudio](www.rstudio.com) environment. Please make sure to have RStudio installed on your computer.

We are going to use the [ggplot2](http://ggplot2.org/) library. If you did not install this package already, install it usng the _Packages_ tab or run `install.packages("ggplot2")`.

```{r libraries}
library(ggplot2)
```

## File import

The MaxQuant report files are tab separated text files that can readily be imported in R as data frame. A data frame allows the convenient manipulation of large tables.

```{r input}
proteinGroupsInput <- read.table(file = "proteinGroups_5cell-line-mix.txt", header = T, stringsAsFactors = F, sep = "\t")
```

You should see the proteinGroupsInput in your _Environment_ panel. You can get the size of the table using the _length_ function.

```{r input_size}
nColumns <- length(proteinGroupsInput)
nLines <- length(proteinGroupsInput$Protein.IDs)
paste("Number of columns: ", nColumns, ", number of protein groups: ", nLines, sep="")
```

## Filtering of the protein groups

MaxQuant indicates contaminants, decoy sequences, and proteins only identified by site by a '+' in their columns. You can access the number of every category by using the _table_ function.

```{r invalid_groups_count}
nContaminants <- table(proteinGroupsInput$Contaminant)
paste("Number of Contaminants: ", nContaminants[2], ", Others: ", nContaminants[1], sep="")
nDecoys <- table(proteinGroupsInput$Reverse)
paste("Number of decoys: ", nDecoys[2], ", Others: ", nDecoys[1], sep="")
nOnlyIdentifiedBySite <- table(proteinGroupsInput$Only.identified.by.site)
paste("Number of only identified by site: ", nOnlyIdentifiedBySite[2], ", Others: ", nOnlyIdentifiedBySite[1], sep="")
```

Filter these lines and put the result in a new table named _proteinGroups.

```{r invalid_groups_filter}
proteinGroups <- proteinGroupsInput[proteinGroupsInput$Contaminant != '+' & proteinGroupsInput$Reverse != '+' & proteinGroupsInput$Only.identified.by.site != '+',]

nLines <- length(proteinGroups$Protein.IDs)
nFiltered <- length(proteinGroupsInput$Protein.IDs) - length(proteinGroups$Protein.IDs)
paste("New number of protein groups: ", nLines, ", Number of protein groups removed: ", nFiltered, sep="")
```

## Extraction of the quantitative columns

The expression values of the five cell lines used are stored in different columns as detailed below. They correspond to two states of the disease, _Diagnosis_ and _Relapse_.

Column | Cell Line | Condition
------ | --------- | ---------
Ratio H/L normalized E1 | Molm-13 | Relapse
Ratio H/L normalized E2 | MV4-11 | Diagnosis
Ratio H/L normalized E3 | NB4 | Relapse
Ratio H/L normalized E4 | OCI-AML3 | Diagnosis
Ratio H/L normalized E5 | THP-1 | Relapse

Note that when importing the data in R, spaces and special characters in the column titles were replaces by dots.

We will now store these values in a new table and save the columns of the cell lines at different conditions in two vectors.

```{r ratios_extraction}
proteinGroupsRatios <- data.frame(proteinIDs = proteinGroups$Protein.IDs, 
                     molm13 = proteinGroups$Ratio.H.L.normalized.E1, 
                     mv411 = proteinGroups$Ratio.H.L.normalized.E2,
                     nb4 = proteinGroups$Ratio.H.L.normalized.E3,
                     ociAml3 = proteinGroups$Ratio.H.L.normalized.E4,
                     thp1 = proteinGroups$Ratio.H.L.normalized.E5)
diagnosisCellLines <- c("mv411", "ociAml3")
relapseCellLines <- c("molm13", "nb4", "thp1")
```

The ratios can be plotted against each others for different cell lines in a scatter plot using the code below.

```{r scatter_plot}
cellLineX <- "mv411"
cellLineY <- "ociAml3"
scatterPlot <- ggplot()
scatterPlot <- scatterPlot + geom_point(aes(x=proteinGroupsRatios[,cellLineX], y=proteinGroupsRatios[,cellLineY]), alpha=0.2, col="blue", size=1)
scatterPlot <- scatterPlot + xlab(cellLineX) + ylab(cellLineY)
plot(scatterPlot)
```

It is possible to extract the ratios of a given protein.

```{r A0FGR8_ratios}
proteinGroupsRatios[proteinGroupsRatios$proteinIDs == "A0FGR8",]
```

It is possible to extract the ratios of a given protein for a given condition.

```{r A0FGR8_diagnosis_ratios}
proteinGroupsRatios[proteinGroupsRatios$proteinIDs == "A0FGR8", names(proteinGroupsRatios) %in% diagnosisCellLines]
```

Note that some proteins have missing values.

```{r A0JLT2_ratios}
proteinGroupsRatios[proteinGroupsRatios$proteinIDs == "A0JLT2",]
```


## Filtering based on the number of missing values

It is possible to see whether a value is missing using the is.na function.

```{r A0JLT2_missing}
is.na(proteinGroupsRatios[proteinGroupsRatios$proteinIDs == "A0JLT2",])
```

The number of missing values for a condition can be obtained with the sum function.

```{r A0JLT2_n_missing}
sum(is.na(proteinGroupsRatios[proteinGroupsRatios$proteinIDs == "A0JLT2", names(proteinGroupsRatios) %in% diagnosisCellLines]))
```

Filter the ratio table to have at least two valid values in every conition, i.e. no missing value at diagnosis and maximum one at relapse. Doing the sum on every row is done using rowSums.

```{r filter_valid_values}
nMissingDiagnosis <- rowSums(is.na(proteinGroupsRatios[, names(proteinGroupsRatios) %in% diagnosisCellLines]))
nMissingRelapse <- rowSums(is.na(proteinGroupsRatios[, names(proteinGroupsRatios) %in% relapseCellLines]))
validProteinGroupsRatios <- proteinGroupsRatios[nMissingDiagnosis == 0 & nMissingRelapse <= 1,]

nLines <- length(validProteinGroupsRatios$proteinIDs)
nFiltered <- length(proteinGroupsRatios$proteinIDs) - length(validProteinGroupsRatios$proteinIDs)
paste("New number of protein groups: ", nLines, ", Number of protein groups removed: ", nFiltered, sep="")
```

The number of protein presenting missing values can be plotted using the code below.

```{r plot_missing_values_histogram}
categoryDiagnosis <- character(length(nMissingDiagnosis))
categoryDiagnosis[] <- "Diagnosis"
categoryRelapse <- character(length(nMissingRelapse))
categoryRelapse[] <- "Relapse"
categories <- c(categoryDiagnosis, categoryRelapse)
values <- c(nMissingDiagnosis, nMissingRelapse)

missingValuesHistogramPlot <- ggplot()
missingValuesHistogramPlot <- missingValuesHistogramPlot + geom_bar(aes(x=values, fill=categories), position = "dodge")
missingValuesHistogramPlot <- missingValuesHistogramPlot + xlab("Number of missing values") + ylab("Density of proteins")
plot(missingValuesHistogramPlot)
```

## Transformation of the ratios

The MaxQuant ratios are estimated as _Heavy/Light_, but our reference was on the _Heavy_ channel, we therefore need to invert the ratios in order to obtain a value representing the cell line to internal standard ratio.

```{r ratio_inversion}
validProteinGroupsRatios[,"molm13"] <- 1/validProteinGroupsRatios[,"molm13"]
validProteinGroupsRatios[,"mv411"] <- 1/validProteinGroupsRatios[,"mv411"]
validProteinGroupsRatios[,"nb4"] <- 1/validProteinGroupsRatios[,"nb4"]
validProteinGroupsRatios[,"ociAml3"] <- 1/validProteinGroupsRatios[,"ociAml3"]
validProteinGroupsRatios[,"thp1"] <- 1/validProteinGroupsRatios[,"thp1"]
```

As one can see, the ratios are distributed over positive values around 1.

```{r ratio_densities}
categories <- c()
categoryCellLine <- character(length(validProteinGroupsRatios[,"molm13"]))
categoryCellLine[] <- "molm13"
categories <- c(categories, categoryCellLine)
categoryCellLine <- character(length(validProteinGroupsRatios[,"mv411"]))
categoryCellLine[] <- "mv411"
categories <- c(categories, categoryCellLine)
categoryCellLine <- character(length(validProteinGroupsRatios[,"nb4"]))
categoryCellLine[] <- "nb4"
categories <- c(categories, categoryCellLine)
categoryCellLine <- character(length(validProteinGroupsRatios[,"ociAml3"]))
categoryCellLine[] <- "ociAml3"
categories <- c(categories, categoryCellLine)
categoryCellLine <- character(length(validProteinGroupsRatios[,"thp1"]))
categoryCellLine[] <- "thp1"
categories <- c(categories, categoryCellLine)
values <- c(validProteinGroupsRatios[,"molm13"],
            validProteinGroupsRatios[,"mv411"],
            validProteinGroupsRatios[,"nb4"],
            validProteinGroupsRatios[,"ociAml3"],
            validProteinGroupsRatios[,"thp1"])
ratiosDensityPlot <- ggplot()
ratiosDensityPlot <- ratiosDensityPlot + geom_density(aes(x=values, col=categories, fill = categories), alpha = 0.1, na.rm=TRUE)
ratiosDensityPlot <- ratiosDensityPlot + xlim(0, 5)
ratiosDensityPlot <- ratiosDensityPlot + xlab("Ratio") + ylab("Density of proteins")
plot(ratiosDensityPlot)
```

Before further processing, we log transform these ratios to restore the symmetry around 1.

```{r ratio_log}
validProteinGroupsRatios[,"molm13"] <- log2(validProteinGroupsRatios[,"molm13"])
validProteinGroupsRatios[,"mv411"] <- log2(validProteinGroupsRatios[,"mv411"])
validProteinGroupsRatios[,"nb4"] <- log2(validProteinGroupsRatios[,"nb4"])
validProteinGroupsRatios[,"ociAml3"] <- log2(validProteinGroupsRatios[,"ociAml3"])
validProteinGroupsRatios[,"thp1"] <- log2(validProteinGroupsRatios[,"thp1"])
```

Now the distributions are centered around 0 and the ratios symmetrically distributed.

```{r ratio_log_densities}
categories <- c()
categoryCellLine <- character(length(validProteinGroupsRatios[,"molm13"]))
categoryCellLine[] <- "molm13"
categories <- c(categories, categoryCellLine)
categoryCellLine <- character(length(validProteinGroupsRatios[,"mv411"]))
categoryCellLine[] <- "mv411"
categories <- c(categories, categoryCellLine)
categoryCellLine <- character(length(validProteinGroupsRatios[,"nb4"]))
categoryCellLine[] <- "nb4"
categories <- c(categories, categoryCellLine)
categoryCellLine <- character(length(validProteinGroupsRatios[,"ociAml3"]))
categoryCellLine[] <- "ociAml3"
categories <- c(categories, categoryCellLine)
categoryCellLine <- character(length(validProteinGroupsRatios[,"thp1"]))
categoryCellLine[] <- "thp1"
categories <- c(categories, categoryCellLine)
values <- c(validProteinGroupsRatios[,"molm13"],
            validProteinGroupsRatios[,"mv411"],
            validProteinGroupsRatios[,"nb4"],
            validProteinGroupsRatios[,"ociAml3"],
            validProteinGroupsRatios[,"thp1"])
ratiosDensityPlot <- ggplot()
ratiosDensityPlot <- ratiosDensityPlot + geom_density(aes(x=values, col=categories, fill = categories), alpha = 0.1, na.rm=TRUE)
ratiosDensityPlot <- ratiosDensityPlot + xlab("Ratio") + ylab("Density of proteins")
plot(ratiosDensityPlot)
```

## Normalization of the ratios

The ratios provided by MaxQuant are already normalized, but it can be useful to conduct an additional normalization to correct for eventual biases. Note: Don't forget to exclude missing values.

```{r ratio_normalization}
cellLineMedian <- median(validProteinGroupsRatios[,"molm13"], na.rm = T)
validProteinGroupsRatios[,"molm13"] <- validProteinGroupsRatios[,"molm13"] - cellLineMedian
cellLineMedian <- median(validProteinGroupsRatios[,"mv411"], na.rm = T)
validProteinGroupsRatios[,"mv411"] <- validProteinGroupsRatios[,"mv411"] - cellLineMedian
cellLineMedian <- median(validProteinGroupsRatios[,"nb4"], na.rm = T)
validProteinGroupsRatios[,"nb4"] <- validProteinGroupsRatios[,"nb4"] - cellLineMedian
cellLineMedian <- median(validProteinGroupsRatios[,"ociAml3"], na.rm = T)
validProteinGroupsRatios[,"ociAml3"] <- validProteinGroupsRatios[,"ociAml3"] - cellLineMedian
cellLineMedian <- median(validProteinGroupsRatios[,"thp1"], na.rm = T)
validProteinGroupsRatios[,"thp1"] <- validProteinGroupsRatios[,"thp1"] - cellLineMedian
```

Removing the median put the center of the distributions at 0.

```{r ratio_log_densities_normalized}
categories <- c()
categoryCellLine <- character(length(validProteinGroupsRatios[,"molm13"]))
categoryCellLine[] <- "molm13"
categories <- c(categories, categoryCellLine)
categoryCellLine <- character(length(validProteinGroupsRatios[,"mv411"]))
categoryCellLine[] <- "mv411"
categories <- c(categories, categoryCellLine)
categoryCellLine <- character(length(validProteinGroupsRatios[,"nb4"]))
categoryCellLine[] <- "nb4"
categories <- c(categories, categoryCellLine)
categoryCellLine <- character(length(validProteinGroupsRatios[,"ociAml3"]))
categoryCellLine[] <- "ociAml3"
categories <- c(categories, categoryCellLine)
categoryCellLine <- character(length(validProteinGroupsRatios[,"thp1"]))
categoryCellLine[] <- "thp1"
categories <- c(categories, categoryCellLine)
values <- c(validProteinGroupsRatios[,"molm13"],
            validProteinGroupsRatios[,"mv411"],
            validProteinGroupsRatios[,"nb4"],
            validProteinGroupsRatios[,"ociAml3"],
            validProteinGroupsRatios[,"thp1"])
ratiosDensityPlot <- ggplot()
ratiosDensityPlot <- ratiosDensityPlot + geom_density(aes(x=values, col=categories, fill = categories), alpha = 0.1, na.rm=TRUE)
ratiosDensityPlot <- ratiosDensityPlot + xlab("Ratio") + ylab("Density of proteins")
plot(ratiosDensityPlot)
```

## t-test

We are going to evaluate the significance of the ratio between the two groups using a t-test. This example does not have sufficient number of replicates to draw any biological conclusion but this is just an example. Below is an example using A0FGR8.

```{r t-test_A0FGR8}
valuesDiagnostic <- validProteinGroupsRatios[validProteinGroupsRatios$proteinIDs == "A0FGR8", names(validProteinGroupsRatios) %in% diagnosisCellLines]
valuesRelapse <- validProteinGroupsRatios[validProteinGroupsRatios$proteinIDs == "A0FGR8", names(validProteinGroupsRatios) %in% relapseCellLines]
t.test(valuesDiagnostic, valuesRelapse, alternative = "two.sided", paired = F)
```

Note that the test run is actually a Welch Two Sample t-test, an extension of the Student t-test more reliable for samples of unequal variances. We create new columns containing the test statistics, as well as the fold change and -10log(p-value).

```{r t-test}
pValues <- c()
ts <- c()
fcs <- c()
pLog <- c()
for (i in 1:length(validProteinGroupsRatios$proteinIDs)) {
  valuesDiagnostic <- validProteinGroupsRatios[i, names(validProteinGroupsRatios) %in% diagnosisCellLines]
  valuesRelapse <- validProteinGroupsRatios[i, names(validProteinGroupsRatios) %in% relapseCellLines]
  test <- t.test(valuesDiagnostic, valuesRelapse, alternative = "two.sided", paired = F)
  pValues <- c(pValues, test$p.value)
  pLog <- c(pLog, -log10(test$p.value))
  ts <- c(ts, test$statistic)
  medianDiagnostic <- median(as.numeric(valuesDiagnostic), na.rm = T)
  medianRelapse <- median(as.numeric(valuesRelapse), na.rm = T)
  fc <- medianRelapse-medianDiagnostic
  fcs <- c(fcs, fc)
}
validProteinGroupsRatios$fc <- fcs
validProteinGroupsRatios$tTestPValue <- pValues
validProteinGroupsRatios$tTestPValueLog <- pLog
validProteinGroupsRatios$tTestStatistic <- ts
```

## QQ Plot

We will now see whether the t-test statistics actually follow a t-distribution. This is achieved by drawing a quantile-quantile plot (qq-plot).

```{r qq_plot}
expectedQuantiles <- rt(length(validProteinGroupsRatios$tTestStatistic), df=5-2)
expectedQuantiles <- sort(expectedQuantiles)
measuredQuantiles <- validProteinGroupsRatios$tTestStatistic
measuredQuantiles <- sort(measuredQuantiles)

qqPlot <- ggplot()
qqPlot <- qqPlot + geom_point(aes(x=expectedQuantiles, y=measuredQuantiles), size = 1, col = "blue")
qqPlot <- qqPlot + geom_line(aes(x=measuredQuantiles, y=measuredQuantiles), size = 1, alpha = 0.5, linetype = "dotted")
qqPlot <- qqPlot + xlab("Expected Quantile") + ylab("Observed Quantile")
plot(qqPlot)
```


## Volcano Plot

We will now plot the p-value against the fold change between the two conditions.

```{r volcano_plot}
volcanoPlot <- ggplot()
volcanoPlot <- volcanoPlot + geom_point(aes(x=validProteinGroupsRatios$fc, y=validProteinGroupsRatios$tTestPValueLog), size = 1, alpha = 0.5, col = "blue")
volcanoPlot <- volcanoPlot + xlab("Fold Change [log2]") + ylab("p-value [-log(p)]")
plot(volcanoPlot)
```


## Missing values imputation

For the following, we are going to impute the missing values and assign them the value 0.

```{r missing_values}
for (i in 1:length(diagnosisCellLines)) {
  validProteinGroupsRatios[,diagnosisCellLines[i]] <- ifelse(is.na(validProteinGroupsRatios[,diagnosisCellLines[i]]), 0, validProteinGroupsRatios[,diagnosisCellLines[i]])
}
for (i in 1:length(relapseCellLines)) {
  validProteinGroupsRatios[,relapseCellLines[i]] <- ifelse(is.na(validProteinGroupsRatios[,relapseCellLines[i]]), 0, validProteinGroupsRatios[,relapseCellLines[i]])
}
```


## Principal component analysis 

We will now draw a principal component analysis (PCA) plot.

```{r pca}

pcaInput <- data.frame(validProteinGroupsRatios$molm13, 
                       validProteinGroupsRatios$mv411, 
                       validProteinGroupsRatios$nb4, 
                       validProteinGroupsRatios$ociAml3, 
                       validProteinGroupsRatios$thp1)
pca <- prcomp(pcaInput)
pc1 <- pca$rotation[,1]
pc2 <- pca$rotation[,2]
totalStd <- sum(pca$sdev)
contribution1 <- round(100*pca$sdev[1]/totalStd)
contribution2 <- round(100*pca$sdev[2]/totalStd)
names <- names(validProteinGroupsRatios)[2:6]

pcaPlot <- ggplot()
pcaPlot <- pcaPlot + geom_point(aes(x=pc1, y=pc2, col=names))
pcaPlot <- pcaPlot + xlab(paste("PC1 [", contribution1, "%]", sep=""))
pcaPlot <- pcaPlot + ylab(paste("PC2 [", contribution2, "%]", sep=""))
plot(pcaPlot)
```




## References

(1) [Vizcaino, J.A. et al., _ProteomeXchange provides globally coordinated proteomics data submission and dissemination_, Nature Biotechnology, 2014](https://www.ncbi.nlm.nih.gov/pubmed/24727771)
(2) [Martens, L. et al., _PRIDE: the proteomics identifications database_, Proteomics, 2005](https://www.ncbi.nlm.nih.gov/pubmed/16041671)
(3) [Geiger, T. et al., _Super-SILAC mix for quantitative proteomics of human tumor tissue_, Nature Methods, 2010](https://www.ncbi.nlm.nih.gov/pubmed/20364148)
(4) [Cox, J. and Mann, M, _MaxQuant enables high peptide identification rates, individualized p.p.b. range mass accuracies and proteome-wide protein quantification_, Nature Biotechnology, 2008](https://www.ncbi.nlm.nih.gov/pubmed/19029910)
(5) [Aasebo, E. et al. Performance of super-SILAC based quantitative proteomics for comparison of
different acute myeloid leukemia (AML) cell lines, Proteomics, 2014](https://www.ncbi.nlm.nih.gov/pubmed/25044641)
