---
title: "Analysis of Quantitative Proteomics Data"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The result of a proteomic quantitative analysis is a list of peptide and protein abundances for every protein in different samples, or abundance ratios between the samples. The downstream interpretation of these data varies according to the experimental design. In this chapter we will describe different generic methods for the interpretation of quantitative datasets.

The dataset used here for illustrative purposes is freely available through the ProteomeXchange [(1)](#references) consortium via the PRIDE [(2)](#references) partner repository under the accession number PXD000441, and available in the _resources_ folder under _Results_MQ_5cell-line-mix_. It consists of five cell lines derived from acute myeloid leukemia (AML) patients measured with a spiked-in internal stanard (IS) obtained from the combination of the same five AML cell lines that have been metabolically labeled with heavy isotopes [(3)](#references) and analyzed using MaxQuant [(4)](#references) version 1.4.1.2. See [(5)](#references) for details.

This chapter introduces the basic methods, and is by no means aiming at covering all possibilities or present a reference workflow. Please continue exploring the data on your own and critically adapt the interpretation workflow to your experiment - there is no one workflow fits all!

## Installation

This tutorial uses [R](https://www.r-project.org), a language that allows the simple manipulation of large datasets. We will use R from the open source [RSudio](www.rstudio.com) environment. Please make sure to have RStudio installed on your computer.

Create a new project via the _File_ -> _New Projet_ menu. Select _New Directory_ -> _Empty Project_, and select a folder for your project. Create a new script via the _File_ -> _New File_ -> _R Script_ menu. 

The commands used in this tutorial are given in line and the results produced are preceded by '##'. You can copy the commands of this tutorial in your script, where one line corresponds to a command. The commands can be run line by line using the _Run_ button. Alternatively, you can run the entire script using the _Source_ button. Note that it is also possible to run the commands in the _Console_.

We are going to use the [ggplot2](http://ggplot2.org/) library. If you did not install this package already, install it usng the _Packages_ tab or run `install.packages("ggplot2")`.

```{r libraries}
library(ggplot2)
```

## File import

The MaxQuant report files are tab separated text files that can readily be imported in R as data frame. A data frame allows the convenient manipulation of large tables. Please download the file [_proteinGroups_5cell-line-mix.txt_](https://github.com/mvaudel/tutorials/blob/master/Expression_data/proteinGroups_5cell-line-mix.txt) to your project folder. It should appear in the _Files_ panel.

```{r input}
proteinGroupsInput <- read.table(file = "proteinGroups_5cell-line-mix.txt", header = T, stringsAsFactors = F, sep = "\t")
```

You should see the proteinGroupsInput in your _Environment_ panel. You can get the size of the table using the _length_ function.

```{r input_size}
nColumns <- length(proteinGroupsInput)
nLines <- length(proteinGroupsInput$Protein.IDs)
paste("Number of columns: ", nColumns, ", number of protein groups: ", nLines, sep="")
```

## Filtering of the protein groups

MaxQuant indicates contaminants, decoy sequences, and proteins only identified by site by '+' in their columns. You can access the number of every category by using the _table_ function.

```{r invalid_groups_count}
nContaminants <- table(proteinGroupsInput$Contaminant)
paste("Number of Contaminants: ", nContaminants[2], ", Others: ", nContaminants[1], sep="")
nDecoys <- table(proteinGroupsInput$Reverse)
paste("Number of decoys: ", nDecoys[2], ", Others: ", nDecoys[1], sep="")
nOnlyIdentifiedBySite <- table(proteinGroupsInput$Only.identified.by.site)
paste("Number of only identified by site: ", nOnlyIdentifiedBySite[2], ", Others: ", nOnlyIdentifiedBySite[1], sep="")
```

Filter these lines and put the result in a new table named _proteinGroups.

```{r invalid_groups_filter}
proteinGroups <- proteinGroupsInput[proteinGroupsInput$Contaminant != '+' & proteinGroupsInput$Reverse != '+' & proteinGroupsInput$Only.identified.by.site != '+',]

nLines <- length(proteinGroups$Protein.IDs)
nFiltered <- length(proteinGroupsInput$Protein.IDs) - length(proteinGroups$Protein.IDs)
paste("New number of protein groups: ", nLines, ", Number of protein groups removed: ", nFiltered, sep="")
```

## Extraction of the quantitative columns

The expression values of the five cell lines used are stored in different columns as detailed below. They correspond to two states of the disease, _Diagnosis_ and _Relapse_.

Column | Cell Line | Condition
------ | --------- | ---------
Ratio H/L normalized E1 | Molm-13 | Relapse
Ratio H/L normalized E2 | MV4-11 | Diagnosis
Ratio H/L normalized E3 | NB4 | Relapse
Ratio H/L normalized E4 | OCI-AML3 | Diagnosis
Ratio H/L normalized E5 | THP-1 | Relapse

Note that when importing the data in R, spaces and special characters in the column titles were replaces by dots.

We will now store these values in a new table and save the columns of the cell lines at different conditions in two vectors.

```{r ratios_extraction}
proteinGroupsRatios <- data.frame(proteinIDs = proteinGroups$Protein.IDs, 
                     molm13 = proteinGroups$Ratio.H.L.normalized.E1, 
                     mv411 = proteinGroups$Ratio.H.L.normalized.E2,
                     nb4 = proteinGroups$Ratio.H.L.normalized.E3,
                     ociAml3 = proteinGroups$Ratio.H.L.normalized.E4,
                     thp1 = proteinGroups$Ratio.H.L.normalized.E5)
diagnosisCellLines <- c("mv411", "ociAml3")
relapseCellLines <- c("molm13", "nb4", "thp1")
```

The ratios can be plotted against each others for different cell lines in a scatter plot using the code below.

```{r scatter_plot}
cellLineX <- "mv411"
cellLineY <- "ociAml3"
scatterPlot <- ggplot()
scatterPlot <- scatterPlot + geom_point(aes(x=proteinGroupsRatios[,cellLineX], y=proteinGroupsRatios[,cellLineY]), alpha=0.2, col="blue", size=1, na.rm = T)
scatterPlot <- scatterPlot + xlab(cellLineX) + ylab(cellLineY)
plot(scatterPlot)
```

It is possible to extract the ratios of a given protein.

```{r A0FGR8_ratios}
proteinGroupsRatios[proteinGroupsRatios$proteinIDs == "A0FGR8",]
```

It is possible to extract the ratios of a given protein for a given condition.

```{r A0FGR8_diagnosis_ratios}
proteinGroupsRatios[proteinGroupsRatios$proteinIDs == "A0FGR8", names(proteinGroupsRatios) %in% diagnosisCellLines]
```

Note that some proteins have missing values.

```{r A0JLT2_ratios}
proteinGroupsRatios[proteinGroupsRatios$proteinIDs == "A0JLT2",]
```


## Filtering based on the number of missing values

It is possible to see whether a value is missing using the is.na function.

```{r A0JLT2_missing}
is.na(proteinGroupsRatios[proteinGroupsRatios$proteinIDs == "A0JLT2",])
```

The number of missing values for a condition can be obtained with the sum function.

```{r A0JLT2_n_missing}
sum(is.na(proteinGroupsRatios[proteinGroupsRatios$proteinIDs == "A0JLT2", names(proteinGroupsRatios) %in% diagnosisCellLines]))
```

Filter the ratio table to have at least two valid values in every conition, i.e. no missing value at diagnosis and maximum one at relapse. Doing the sum on every row is done using rowSums.

```{r filter_valid_values}
nMissingDiagnosis <- rowSums(is.na(proteinGroupsRatios[, names(proteinGroupsRatios) %in% diagnosisCellLines]))
nMissingRelapse <- rowSums(is.na(proteinGroupsRatios[, names(proteinGroupsRatios) %in% relapseCellLines]))
validProteinGroupsRatios <- proteinGroupsRatios[nMissingDiagnosis == 0 & nMissingRelapse <= 1,]

nLines <- length(validProteinGroupsRatios$proteinIDs)
nFiltered <- length(proteinGroupsRatios$proteinIDs) - length(validProteinGroupsRatios$proteinIDs)
paste("New number of protein groups: ", nLines, ", Number of protein groups removed: ", nFiltered, sep="")
```

The number of protein presenting missing values can be plotted using the code below.

```{r plot_missing_values_histogram}
categoryDiagnosis <- character(length(nMissingDiagnosis))
categoryDiagnosis[] <- "Diagnosis"
categoryRelapse <- character(length(nMissingRelapse))
categoryRelapse[] <- "Relapse"
categories <- c(categoryDiagnosis, categoryRelapse)
values <- c(nMissingDiagnosis, nMissingRelapse)

missingValuesHistogramPlot <- ggplot()
missingValuesHistogramPlot <- missingValuesHistogramPlot + geom_bar(aes(x=values, fill=categories), position = "dodge")
missingValuesHistogramPlot <- missingValuesHistogramPlot + xlab("Number of missing values") + ylab("Number of proteins")
plot(missingValuesHistogramPlot)
```

## Transformation of the ratios

The MaxQuant ratios are estimated as _Heavy/Light_, but our reference was on the _Heavy_ channel, we therefore need to invert the ratios in order to obtain a value representing the cell line to internal standard ratio.

```{r ratio_inversion}
validProteinGroupsRatios[,"molm13"] <- 1/validProteinGroupsRatios[,"molm13"]
validProteinGroupsRatios[,"mv411"] <- 1/validProteinGroupsRatios[,"mv411"]
validProteinGroupsRatios[,"nb4"] <- 1/validProteinGroupsRatios[,"nb4"]
validProteinGroupsRatios[,"ociAml3"] <- 1/validProteinGroupsRatios[,"ociAml3"]
validProteinGroupsRatios[,"thp1"] <- 1/validProteinGroupsRatios[,"thp1"]
```

As one can see, the ratios are distributed over positive values around 1.

```{r ratio_densities}
categories <- c()
categoryCellLine <- character(length(validProteinGroupsRatios[,"molm13"]))
categoryCellLine[] <- "molm13"
categories <- c(categories, categoryCellLine)
categoryCellLine <- character(length(validProteinGroupsRatios[,"mv411"]))
categoryCellLine[] <- "mv411"
categories <- c(categories, categoryCellLine)
categoryCellLine <- character(length(validProteinGroupsRatios[,"nb4"]))
categoryCellLine[] <- "nb4"
categories <- c(categories, categoryCellLine)
categoryCellLine <- character(length(validProteinGroupsRatios[,"ociAml3"]))
categoryCellLine[] <- "ociAml3"
categories <- c(categories, categoryCellLine)
categoryCellLine <- character(length(validProteinGroupsRatios[,"thp1"]))
categoryCellLine[] <- "thp1"
categories <- c(categories, categoryCellLine)
values <- c(validProteinGroupsRatios[,"molm13"],
            validProteinGroupsRatios[,"mv411"],
            validProteinGroupsRatios[,"nb4"],
            validProteinGroupsRatios[,"ociAml3"],
            validProteinGroupsRatios[,"thp1"])
ratiosDensityPlot <- ggplot()
ratiosDensityPlot <- ratiosDensityPlot + geom_density(aes(x=values, col=categories, fill = categories), alpha = 0.1, na.rm=TRUE)
ratiosDensityPlot <- ratiosDensityPlot + xlim(0, 5)
ratiosDensityPlot <- ratiosDensityPlot + xlab("Ratio") + ylab("Density of proteins")
plot(ratiosDensityPlot)
```

Before further processing, we log transform these ratios to restore the symmetry around 1.

```{r ratio_log}
validProteinGroupsRatios[,"molm13"] <- log2(validProteinGroupsRatios[,"molm13"])
validProteinGroupsRatios[,"mv411"] <- log2(validProteinGroupsRatios[,"mv411"])
validProteinGroupsRatios[,"nb4"] <- log2(validProteinGroupsRatios[,"nb4"])
validProteinGroupsRatios[,"ociAml3"] <- log2(validProteinGroupsRatios[,"ociAml3"])
validProteinGroupsRatios[,"thp1"] <- log2(validProteinGroupsRatios[,"thp1"])
```

Now the distributions are centered around 0 and the ratios symmetrically distributed.

```{r ratio_log_densities}
categories <- c()
categoryCellLine <- character(length(validProteinGroupsRatios[,"molm13"]))
categoryCellLine[] <- "molm13"
categories <- c(categories, categoryCellLine)
categoryCellLine <- character(length(validProteinGroupsRatios[,"mv411"]))
categoryCellLine[] <- "mv411"
categories <- c(categories, categoryCellLine)
categoryCellLine <- character(length(validProteinGroupsRatios[,"nb4"]))
categoryCellLine[] <- "nb4"
categories <- c(categories, categoryCellLine)
categoryCellLine <- character(length(validProteinGroupsRatios[,"ociAml3"]))
categoryCellLine[] <- "ociAml3"
categories <- c(categories, categoryCellLine)
categoryCellLine <- character(length(validProteinGroupsRatios[,"thp1"]))
categoryCellLine[] <- "thp1"
categories <- c(categories, categoryCellLine)
values <- c(validProteinGroupsRatios[,"molm13"],
            validProteinGroupsRatios[,"mv411"],
            validProteinGroupsRatios[,"nb4"],
            validProteinGroupsRatios[,"ociAml3"],
            validProteinGroupsRatios[,"thp1"])
ratiosDensityPlot <- ggplot()
ratiosDensityPlot <- ratiosDensityPlot + geom_density(aes(x=values, col=categories, fill = categories), alpha = 0.1, na.rm=TRUE)
ratiosDensityPlot <- ratiosDensityPlot + xlab("Ratio") + ylab("Density of proteins")
plot(ratiosDensityPlot)
```

## Normalization of the ratios

The ratios provided by MaxQuant are already normalized, but it can be useful to conduct an additional normalization to correct for eventual biases. Note: Don't forget to exclude missing values.

```{r ratio_normalization}
cellLineMedian <- median(validProteinGroupsRatios[,"molm13"], na.rm = T)
validProteinGroupsRatios[,"molm13"] <- validProteinGroupsRatios[,"molm13"] - cellLineMedian
cellLineMedian <- median(validProteinGroupsRatios[,"mv411"], na.rm = T)
validProteinGroupsRatios[,"mv411"] <- validProteinGroupsRatios[,"mv411"] - cellLineMedian
cellLineMedian <- median(validProteinGroupsRatios[,"nb4"], na.rm = T)
validProteinGroupsRatios[,"nb4"] <- validProteinGroupsRatios[,"nb4"] - cellLineMedian
cellLineMedian <- median(validProteinGroupsRatios[,"ociAml3"], na.rm = T)
validProteinGroupsRatios[,"ociAml3"] <- validProteinGroupsRatios[,"ociAml3"] - cellLineMedian
cellLineMedian <- median(validProteinGroupsRatios[,"thp1"], na.rm = T)
validProteinGroupsRatios[,"thp1"] <- validProteinGroupsRatios[,"thp1"] - cellLineMedian
```

Removing the median put the center of the distributions at 0.

```{r ratio_log_densities_normalized}
categories <- c()
categoryCellLine <- character(length(validProteinGroupsRatios[,"molm13"]))
categoryCellLine[] <- "molm13"
categories <- c(categories, categoryCellLine)
categoryCellLine <- character(length(validProteinGroupsRatios[,"mv411"]))
categoryCellLine[] <- "mv411"
categories <- c(categories, categoryCellLine)
categoryCellLine <- character(length(validProteinGroupsRatios[,"nb4"]))
categoryCellLine[] <- "nb4"
categories <- c(categories, categoryCellLine)
categoryCellLine <- character(length(validProteinGroupsRatios[,"ociAml3"]))
categoryCellLine[] <- "ociAml3"
categories <- c(categories, categoryCellLine)
categoryCellLine <- character(length(validProteinGroupsRatios[,"thp1"]))
categoryCellLine[] <- "thp1"
categories <- c(categories, categoryCellLine)
values <- c(validProteinGroupsRatios[,"molm13"],
            validProteinGroupsRatios[,"mv411"],
            validProteinGroupsRatios[,"nb4"],
            validProteinGroupsRatios[,"ociAml3"],
            validProteinGroupsRatios[,"thp1"])
ratiosDensityPlot <- ggplot()
ratiosDensityPlot <- ratiosDensityPlot + geom_density(aes(x=values, col=categories, fill = categories), alpha = 0.1, na.rm=TRUE)
ratiosDensityPlot <- ratiosDensityPlot + xlab("Ratio") + ylab("Density of proteins")
plot(ratiosDensityPlot)
```

## t-test

We are going to evaluate the significance of the ratio between the two groups using a t-test. This example does not have sufficient number of replicates to draw any biological conclusion but this is just an example. Below is an example using A0FGR8.

```{r t-test_A0FGR8}
valuesDiagnostic <- validProteinGroupsRatios[validProteinGroupsRatios$proteinIDs == "A0FGR8", names(validProteinGroupsRatios) %in% diagnosisCellLines]
valuesRelapse <- validProteinGroupsRatios[validProteinGroupsRatios$proteinIDs == "A0FGR8", names(validProteinGroupsRatios) %in% relapseCellLines]
t.test(valuesDiagnostic, valuesRelapse, alternative = "two.sided", paired = F)
```

Note that the test run is actually a Welch Two Sample t-test, an extension of the Student t-test more reliable for samples of unequal variances. We create new columns containing the test statistics, as well as the fold change and -10log(p-value).

```{r t-test}
pValues <- c()
ts <- c()
fcs <- c()
pLog <- c()
for (i in 1:length(validProteinGroupsRatios$proteinIDs)) {
  valuesDiagnostic <- validProteinGroupsRatios[i, names(validProteinGroupsRatios) %in% diagnosisCellLines]
  valuesRelapse <- validProteinGroupsRatios[i, names(validProteinGroupsRatios) %in% relapseCellLines]
  test <- t.test(valuesDiagnostic, valuesRelapse, alternative = "two.sided", paired = F)
  pValues <- c(pValues, test$p.value)
  pLog <- c(pLog, -log10(test$p.value))
  ts <- c(ts, test$statistic)
  medianDiagnostic <- median(as.numeric(valuesDiagnostic), na.rm = T)
  medianRelapse <- median(as.numeric(valuesRelapse), na.rm = T)
  fc <- medianRelapse-medianDiagnostic
  fcs <- c(fcs, fc)
}
validProteinGroupsRatios$fc <- fcs
validProteinGroupsRatios$tTestPValue <- pValues
validProteinGroupsRatios$tTestPValueLog <- pLog
validProteinGroupsRatios$tTestStatistic <- ts
```

## QQ Plot

We will now see whether the t-test statistics actually follow a t-distribution. This is achieved by drawing a quantile-quantile plot (qq-plot) where the quantile of the observed t-statistics is plotted against the theoretical quantiles of the Student t distribution.

```{r qq_plot}

degreesOfFreedom <- 5-2 # 5 cell lines

expectedQuantiles <- rt(length(validProteinGroupsRatios$tTestStatistic), df=degreesOfFreedom)
expectedQuantiles <- sort(expectedQuantiles)
measuredQuantiles <- validProteinGroupsRatios$tTestStatistic
measuredQuantiles <- sort(measuredQuantiles)

qqPlot <- ggplot()
qqPlot <- qqPlot + geom_point(aes(x=expectedQuantiles, y=measuredQuantiles), size = 1, col = "blue")
qqPlot <- qqPlot + geom_line(aes(x=expectedQuantiles, y=expectedQuantiles), size = 1, alpha = 0.5, linetype = "dotted")
qqPlot <- qqPlot + xlab("Expected Quantile") + ylab("Observed Quantile")
plot(qqPlot)
```

It is possible to make the same plot for the observed and expected p-values, based on these quantiles. The red dots show a tolerated deviation ratio of from the expected p-values as set in the _ppTolerance_ variable. Note that the measured p-values deviate from the diagonal at low p-values, indicating that our dataset does not fully satisfy the hypotheses of the t-test.

```{r pp_plot}

ppTolerance <-1.1

expectedPValues <- pt(expectedQuantiles, lower.tail = expectedQuantiles < 0, df=degreesOfFreedom)
expectedPValues <- sort(expectedPValues)
expectedPValuesLog <- -log10(expectedPValues)
measuredPValues <- validProteinGroupsRatios$tTestPValue
measuredPValues <- sort(measuredPValues)
measuredPValuesLog <- -log10(measuredPValues)

limitLow <- expectedPValues/ppTolerance
limitLow <- -log10(limitLow)
limitHigh <- ppTolerance * expectedPValues
limitHigh <- -log10(limitHigh)

ppPlot <- ggplot()
ppPlot <- ppPlot + geom_point(aes(x=expectedPValuesLog, y=measuredPValuesLog), size = 1, col = "blue")
ppPlot <- ppPlot + geom_line(aes(x=expectedPValuesLog, y=expectedPValuesLog), size = 1, alpha = 0.5, linetype = "dashed", col="black")
ppPlot <- ppPlot + geom_line(aes(x=expectedPValuesLog, y=limitLow), size = 1, alpha = 0.5, linetype = "dotted", col="red")
ppPlot <- ppPlot + geom_line(aes(x=expectedPValuesLog, y=limitHigh), size = 1, alpha = 0.5, linetype = "dotted", col="red")
ppPlot <- ppPlot + xlab("Expected p-value [-log10(p)]") + ylab("Observed p-value [-log10(p)]")
plot(ppPlot)
```

We calculate lambda as the ratio of the observed p-value and the expected p-value. A density plot shows that the majority of the p-values are within a `r ppTolerance` deviation ratio.

```{r hist_lambda}

lambda <- measuredPValues/expectedPValues
lambdaLog <- log(lambda, base = ppTolerance)

lambdaHistogramPlot <- ggplot()
lambdaHistogramPlot <- lambdaHistogramPlot + geom_density(aes(x=lambdaLog), col="blue", fill = "blue", alpha = 0.1)
lambdaHistogramPlot <- lambdaHistogramPlot + geom_vline(aes(xintercept=-1), col="red", alpha = 0.8, linetype="dotted")
lambdaHistogramPlot <- lambdaHistogramPlot + geom_vline(aes(xintercept=1), col="red", alpha = 0.8, linetype="dotted")
lambdaHistogramPlot <- lambdaHistogramPlot + xlab(paste("Lambda (Observed p-value / Expected p-value) [log", ppTolerance, "]", sep="")) + ylab("Density of proteins")
plot(lambdaHistogramPlot)

corePopulation <- round(100*length(lambdaLog[abs(lambdaLog) < 1])/length(lambdaLog))/100
```

`r 100*corePopulation`% of the p-values is contained within the `r ppTolerance` deviation ratio. This percentile will be used in the following as core population.

## Multiple hypothesis testing

The more we run t-tests, the more chances we have to get a low p-value by chance. For example, when we run the test just one time, we are unlikely to find a p < 1% by chance, whereas if we run the test 100 times, we are likely to find at least one p < 1% by chance. This problem is called the multiple hypothesis testing. Correction for multiple hypothesis testing approaches were pioneered by Benjamini and Hochberg [(6)](#references). Below is an attempt at illustrating this problem intuitively. 

In the following we derive the number of randomly occurring low p-values, false discoveries (FP), from the t-distribution calibrated to the core population. The obtained distribution of false positives is displayed in red. Note that the density of the t-distribution inferred from the core population is used here instead of the p-value, the code can be adapted to use the p-value according to your preferences.

```{r core_population}

tScoreLimits <- qt(c(0.5-corePopulation/2, 0.5+corePopulation/2), df = degreesOfFreedom)
tPopulation <- length(validProteinGroupsRatios$tTestStatistic[validProteinGroupsRatios$tTestStatistic > tScoreLimits[1] & validProteinGroupsRatios$tTestStatistic < tScoreLimits[2]])/corePopulation

nBinsInOne <- 5
binSize <- 1/nBinsInOne
minStatistic <- floor(min(validProteinGroupsRatios$tTestStatistic))
maxStatistic <- ceiling(max(validProteinGroupsRatios$tTestStatistic))

quantilesStatistic <- quantile(validProteinGroupsRatios$tTestStatistic, c(0.5-corePopulation/2, 0.5, 0.5+corePopulation/2), na.rm = T, names = F)
medianStatistic <- quantilesStatistic[2]
interQuantilesStatistic <- quantilesStatistic[3] - quantilesStatistic[1]
xDistribution <- minStatistic + (binSize * (1:(nBinsInOne * (maxStatistic - minStatistic))))
yDistributionT <- dt(x = xDistribution, df = degreesOfFreedom)
scale <- tPopulation / nBinsInOne
yDistributionTNorm <- scale * yDistributionT

tStatisticHistogramPlot <- ggplot()
tStatisticHistogramPlot <- tStatisticHistogramPlot + geom_histogram(aes(x=validProteinGroupsRatios$tTestStatistic), col="blue", fill = "blue", alpha = 0.1, binwidth = binSize)
tStatisticHistogramPlot <- tStatisticHistogramPlot + geom_line(aes(x=xDistribution, yDistributionTNorm), col="red", alpha = 0.8)
tStatisticHistogramPlot <- tStatisticHistogramPlot + xlab("t-statistic") + ylab("Number of proteins")
plot(tStatisticHistogramPlot)
```

By dividing the estimated number of false positives by the number of proteins at a given t-statistic at increasing p-value, we obtain a false discovery rate, _FDR_, the share of proteins incorrectly marked as differentially expressed between the cell lines. In the following, we estimate the FDR at every p-value and plot it against the (log transformed) p-value.

Since the number of hits can fluctuate around the estimated number of false positives, the FDR evolves in steps that can go over 100%. The lowers FDR achievable at a given p-value is therefore used instead. This is termed a _q-value_.

```{r fdr_plot}

fdr <- c()
index <- 1
for (i in order(validProteinGroupsRatios$tTestPValue)) {
  tStatistic <- validProteinGroupsRatios$tTestStatistic[i]
  nFpAtI <- tPopulation * 2 * pt(tStatistic, df = degreesOfFreedom, lower.tail = tStatistic < 0)
  fdrAtI <- nFpAtI / index
  fdr[i] <- fdrAtI
  index <- index + 1
}
validProteinGroupsRatios$fdr <- fdr
fdrPercent <- 100 * fdr

qValue <- c()
lastLowesQValue <- 1
pValue5 <- NA
pValue10 <- NA
for (i in order(validProteinGroupsRatios$tTestPValue, decreasing = T)) {
  fdrAtI <- validProteinGroupsRatios$fdr[i]
  if (fdrAtI < lastLowesQValue) {
    lastLowesQValue <- fdrAtI
  }
  qValue[i] <- lastLowesQValue
  if (is.na(pValue10) && lastLowesQValue < 0.1) {
    pValue10 <- validProteinGroupsRatios$tTestPValueLog[i]
  }
  if (is.na(pValue5) && lastLowesQValue < 0.05) {
    pValue5 <- validProteinGroupsRatios$tTestPValueLog[i]
  }
}
if (is.na(pValue10)) {
  pValue10 <- validProteinGroupsRatios$tTestPValueLog[i]
}
if (is.na(pValue5)) {
  pValue5 <- validProteinGroupsRatios$tTestPValueLog[i]
}
validProteinGroupsRatios$qValue <- qValue
qValuePercent <- 100 * qValue
fdrPlot <- ggplot()
fdrPlot <- fdrPlot + geom_point(aes(x=validProteinGroupsRatios$tTestPValueLog, y=fdrPercent), alpha=0.5, size = 1, col = "blue")
fdrPlot <- fdrPlot + geom_line(aes(x=validProteinGroupsRatios$tTestPValueLog, y=qValuePercent), col = "red")
fdrPlot <- fdrPlot + geom_vline(aes(xintercept = pValue5), col="darkgreen", alpha = 0.8, linetype="dotted", size = 1)
fdrPlot <- fdrPlot + geom_vline(aes(xintercept = pValue10), col="darkorange", alpha = 0.8, linetype="dotted", size = 1)
fdrPlot <- fdrPlot + xlab("p-value") + ylab("FDR (%)")
plot(fdrPlot)
```

In this example, 10% FDR is achieved at p-value `r round(10^-pValue10, 3)` and 5% FDR at p-value `r round(10^-pValue5, 3)`, as indicated by the orange and green lines, respectively. the lowest q-value is `r round(100*lastLowesQValue, 2)`%.


## Independent weighting hypothesis

The following code shows the histogram of fold changes between diagnostic and relapse for all proteins, as estimated in the t-test section by the difference between the median value of relapse cell lines compaired to the median value of diagnostic cell lines. The red line shows a normal distribution with mean and standard deviation calibrated on the median and the 34th percentiles around the median, respectively.

```{r fc_histogram}
nBinsInOne <- 10
binSize <- 1/nBinsInOne
minFC <- floor(min(validProteinGroupsRatios$fc))
maxFC <- ceiling(max(validProteinGroupsRatios$fc))

quantilesFC <- quantile(validProteinGroupsRatios$fc, c(0.16, 0.5, 0.84), na.rm = T, names = F)
medianFC <- quantilesFC[2]
interQuantilesFC <- quantilesFC[3] - quantilesFC[1]
xDistribution <- minFC + (binSize * (1:(nBinsInOne * (maxFC - minFC))))
yDistribution <- dnorm(x = xDistribution, mean = medianFC, sd = interQuantilesFC/2)
scale <- length(validProteinGroupsRatios$fc)/nBinsInOne
yDistributionNorm <- scale * yDistribution

fcHistogramPlot <- ggplot()
fcHistogramPlot <- fcHistogramPlot + geom_histogram(aes(x=validProteinGroupsRatios$fc), col="blue", fill = "blue", alpha = 0.1, binwidth = binSize)
fcHistogramPlot <- fcHistogramPlot + geom_line(aes(x=xDistribution, yDistributionNorm), col="red", alpha = 0.8)
fcHistogramPlot <- fcHistogramPlot + xlab("Fold Change") + ylab("Number of proteins")
plot(fcHistogramPlot)
```

With the hypothesis that false positives introduced by multiple hypothesis testing are likely to distribute according to the rest of the fold changes, independently of the p-value, we can use the fold change as independent weighting hypothesis (IWH) for the FDR estimation. In the following, we sort the proteins according to their probability to belong to the fold change and student distributions. The q-value estimated previously is plotted in blue, and the q-value after IWH is colored in green, orange, and red for q-values below 1%, between 1 and 5%, and higher than 5%, respectively.

```{r fdr_correction_plot}

scoreNorm <- dnorm(validProteinGroupsRatios$fc, mean = medianFC, sd = interQuantilesFC/2)
scoreT <- dt(validProteinGroupsRatios$tTestStatistic, df =degreesOfFreedom)
tNormScore <- scoreNorm * scoreT
validProteinGroupsRatios$tNormScore <- tNormScore
scoreSums <- c()
currentSum <- 0
maxIncrease <- 1/tPopulation # Cannot be more than one FP per protein
for (i in order(validProteinGroupsRatios$tNormScore)) {
  increase <- tNormScore[i]
  if (increase > maxIncrease) {
    increase <- maxIncrease
  }
  currentSum <- currentSum + increase
  scoreSums[i] <- currentSum
}
scoreSums <- scoreSums / currentSum * tPopulation
correctedFdr <- c()
index <- 1
for (i in order(validProteinGroupsRatios$tNormScore)) {
  correctedFdr[i] <- scoreSums[i] / index
  index <- index + 1
}
validProteinGroupsRatios$correctedFdr <- correctedFdr
correctedQValue <- c()
lastLowesQValue <- 1
for (i in order(validProteinGroupsRatios$tNormScore, decreasing = T)) {
  fdrAtI <- validProteinGroupsRatios$correctedFdr[i]
  if (fdrAtI < lastLowesQValue) {
    lastLowesQValue <- fdrAtI
  }
  correctedQValue[i] <- lastLowesQValue
}
validProteinGroupsRatios$correctedQValue <- correctedQValue

correctedFdrPercent <- 100 * correctedFdr
correctedQValuePercent <- 100 * correctedQValue

regulationConfidence <- ifelse(correctedQValuePercent <= 5, "Middle", "Low")
regulationConfidence <- ifelse(correctedQValuePercent <= 1, "High", regulationConfidence)

fdrPlot <- ggplot()
fdrPlot <- fdrPlot + geom_point(aes(x=tNormScore, y=correctedFdrPercent), alpha=0.5, size = 1, col = "blue")
fdrPlot <- fdrPlot + geom_line(aes(x=tNormScore, y=correctedQValuePercent), col = "red")
fdrPlot <- fdrPlot + xlab("p(t)*p(norm)") + ylab("FDR (%)")
plot(fdrPlot)

correctedFdrPlot <- ggplot()
correctedFdrPlot <- correctedFdrPlot + geom_point(aes(x=tNormScore, y=qValuePercent), size = 1, alpha=0.5, col = "blue")
correctedFdrPlot <- correctedFdrPlot + geom_point(aes(x=tNormScore, y=correctedQValuePercent, col=regulationConfidence), size = 1, alpha=0.5)
correctedFdrPlot <- correctedFdrPlot + scale_color_manual(values=c("darkgreen", "darkRed", "darkorange"), name="Regulation Confidence", breaks=c("High", "Middle", "Low"), labels=c("High", "Middle", "Low"))
correctedFdrPlot <- correctedFdrPlot + xlab("p(t)*p(norm)") + ylab("FDR (%)")
plot(correctedFdrPlot)
```

In the following, we plot the p-value against the fold change between the two conditions. This plot is called a _volcano plot_. The protein is colored depending on its q-value as detailed above.

```{r volcano_plot}
volcanoPlot <- ggplot()
volcanoPlot <- volcanoPlot + geom_point(aes(x=validProteinGroupsRatios$fc, y=validProteinGroupsRatios$tTestPValueLog, col=regulationConfidence), size = 1, alpha = 0.5)
volcanoPlot <- volcanoPlot + scale_color_manual(values=c("darkgreen", "darkRed", "darkorange"), name="Regulation Confidence", breaks=c("High", "Middle", "Low"), labels=c("High", "Middle", "Low"))
volcanoPlot <- volcanoPlot + xlab("Fold Change [log2]") + ylab("p-value [-log(p)]")
plot(volcanoPlot)
```

Note that this can easily be combined with an arbirary p-value threshold, e.g. 0.05.

```{r setPLimit}
pLimit <- 0.05
```

For example, in the section below, we conduct the same analysis for proteins with p-value < 0.05.

```{r fdr_correction_0.05}

scoreNorm <- dnorm(validProteinGroupsRatios$fc, mean = medianFC, sd = interQuantilesFC/2)
scoreT <- dt(validProteinGroupsRatios$tTestStatistic, df =degreesOfFreedom)
tNormScore05 <- ifelse(validProteinGroupsRatios$tTestPValue < pLimit, scoreNorm * scoreT, 1)
validProteinGroupsRatios$tNormScore05 <- tNormScore05
scoreSums05 <- c()
currentSum <- 0
maxIncrease <- 1/tPopulation # Cannot be more than one FP per protein
for (i in order(validProteinGroupsRatios$tNormScore05)) {
  increase <- tNormScore05[i]
  if (increase > maxIncrease) {
    increase <- maxIncrease
  }
  currentSum <- currentSum + increase
  scoreSums05[i] <- currentSum
}
scoreSums05 <- scoreSums05 / currentSum * tPopulation
correctedFdr05 <- c()
index <- 1
for (i in order(validProteinGroupsRatios$tNormScore05)) {
  correctedFdr05[i] <- scoreSums05[i] / index
  index <- index + 1
}
validProteinGroupsRatios$correctedFdr05 <- correctedFdr05
correctedQValue05 <- c()
lastLowesQValue <- 1
for (i in order(validProteinGroupsRatios$tNormScore05, decreasing = T)) {
  fdrAtI <- validProteinGroupsRatios$correctedFdr05[i]
  if (fdrAtI < lastLowesQValue) {
    lastLowesQValue <- fdrAtI
  }
  correctedQValue05[i] <- lastLowesQValue
}
validProteinGroupsRatios$correctedQValue05 <- correctedQValue05

correctedFdr05Percent <- 100 * correctedFdr05
correctedQValue05Percent <- 100 * correctedQValue05

regulationConfidence05 <- ifelse(correctedQValue05Percent <= 5, "Middle", "Low")
regulationConfidence05 <- ifelse(correctedQValue05Percent <= 1, "High", regulationConfidence05)

fdrPlot05 <- ggplot()
fdrPlot05 <- fdrPlot05 + geom_point(aes(x=tNormScore05, y=correctedFdr05Percent), alpha=0.5, size = 1, col = "blue")
fdrPlot05 <- fdrPlot05 + geom_line(aes(x=tNormScore05, y=correctedQValue05Percent), col = "red")
fdrPlot05 <- fdrPlot05 + xlab("p(t)*p(norm)") + ylab("FDR (%)")
plot(fdrPlot05)

correctedFdrPlot05 <- ggplot()
correctedFdrPlot05 <- correctedFdrPlot05 + geom_point(aes(x=tNormScore05, y=qValuePercent), size = 1, alpha=0.5, col = "blue")
correctedFdrPlot05 <- correctedFdrPlot05 + geom_point(aes(x=tNormScore05, y=correctedQValue05Percent, col=regulationConfidence05), size = 1, alpha=0.5)
correctedFdrPlot05 <- correctedFdrPlot05 + scale_color_manual(values=c("darkgreen", "darkRed", "darkorange"), name="Regulation Confidence", breaks=c("High", "Middle", "Low"), labels=c("High", "Middle", "Low"))
correctedFdrPlot05 <- correctedFdrPlot05 + xlab("p(t)*p(norm)") + ylab("FDR (%)")
plot(correctedFdrPlot05)

volcanoPlot05 <- ggplot()
volcanoPlot05 <- volcanoPlot05 + geom_point(aes(x=validProteinGroupsRatios$fc, y=validProteinGroupsRatios$tTestPValueLog, col=regulationConfidence05), size = 1, alpha = 0.5)
volcanoPlot05 <- volcanoPlot05 + scale_color_manual(values=c("darkgreen", "darkRed", "darkorange"), name="Regulation Confidence", breaks=c("High", "Middle", "Low"), labels=c("High", "Middle", "Low"))
volcanoPlot05 <- volcanoPlot05 + geom_hline(aes(yintercept = -log10(pLimit)), col="blue", alpha = 0.8, linetype="dotted", size = 1)
volcanoPlot05 <- volcanoPlot05 + xlab("Fold Change [log2]") + ylab("p-value [-log(p)]")
plot(volcanoPlot05)
```


## Missing values imputation

For the following, we are going to impute the missing values and assign them the value 0.

```{r missing_values}
for (i in 1:length(diagnosisCellLines)) {
  validProteinGroupsRatios[,diagnosisCellLines[i]] <- ifelse(is.na(validProteinGroupsRatios[,diagnosisCellLines[i]]), 0, validProteinGroupsRatios[,diagnosisCellLines[i]])
}
for (i in 1:length(relapseCellLines)) {
  validProteinGroupsRatios[,relapseCellLines[i]] <- ifelse(is.na(validProteinGroupsRatios[,relapseCellLines[i]]), 0, validProteinGroupsRatios[,relapseCellLines[i]])
}
```


## Principal component analysis 

We will now draw a principal component analysis (PCA) plot.

```{r pca}

pcaInput <- data.frame(validProteinGroupsRatios$molm13, 
                       validProteinGroupsRatios$mv411, 
                       validProteinGroupsRatios$nb4, 
                       validProteinGroupsRatios$ociAml3, 
                       validProteinGroupsRatios$thp1)
pca <- prcomp(pcaInput)
pc1 <- pca$rotation[,1]
pc2 <- pca$rotation[,2]
totalStd <- sum(pca$sdev)
contribution1 <- round(100*pca$sdev[1]/totalStd)
contribution2 <- round(100*pca$sdev[2]/totalStd)
names <- names(validProteinGroupsRatios)[2:6]

pcaPlot <- ggplot()
pcaPlot <- pcaPlot + geom_point(aes(x=pc1, y=pc2, col=names))
pcaPlot <- pcaPlot + xlab(paste("PC1 [", contribution1, "%]", sep=""))
pcaPlot <- pcaPlot + ylab(paste("PC2 [", contribution2, "%]", sep=""))
plot(pcaPlot)
```




## References

(1) [Vizcaino, J.A. _et al._, _ProteomeXchange provides globally coordinated proteomics data submission and dissemination_, Nature Biotechnology, 2014](https://www.ncbi.nlm.nih.gov/pubmed/24727771)
(2) [Martens, L. _et al._, _PRIDE: the proteomics identifications database_, Proteomics, 2005](https://www.ncbi.nlm.nih.gov/pubmed/16041671)
(3) [Geiger, T. _et al._, _Super-SILAC mix for quantitative proteomics of human tumor tissue_, Nature Methods, 2010](https://www.ncbi.nlm.nih.gov/pubmed/20364148)
(4) [Cox, J. and Mann, M, _MaxQuant enables high peptide identification rates, individualized p.p.b. range mass accuracies and proteome-wide protein quantification_, Nature Biotechnology, 2008](https://www.ncbi.nlm.nih.gov/pubmed/19029910)
(5) [Aasebo, E. _et al._, _Performance of super-SILAC based quantitative proteomics for comparison of
different acute myeloid leukemia (AML) cell lines_, Proteomics, 2014](https://www.ncbi.nlm.nih.gov/pubmed/25044641)
(6) [Benjamini, Y. and Hochberg, Y., _Controlling the false discovery rate: a practical and powerful approach to multiple testing_, Journal of the Royal Statistical Society, 1995](http://www.math.tau.ac.il/~ybenja/MyPapers/benjamini_hochberg1995.pdf)
